Bow shocks, bow waves, and dust waves

### 2018-05-28 papers/Papers1234/sec-dust-bow-intro.tex
Write about rip point

As \(\abs{\phi}\) increases, long-range electrostatic interactions with
protons within the Debye radius (Coulomb drag) become increasingly
important at subsonic relative velocities, as shown in
Figure~\ref{fig:gas-grain-drag-photoionized} by the orange
(\(\phi = 1\)), green (\(\phi = 4\)), and red (\(\phi = 16\)) lines.  However,
the Coulomb drag has a peak when \(w\) is equal to the thermal
speed of the colliders, which is \(\approx \SI{10}{km.s^{-1}}\) for protons,
giving a maximum strength of
\begin{equation}
  \label{eq:fdrag-maximum}
  f_{\mathrm{max}} = 0.5\, (\ln\Lambda)\, \phi^2 f_* \approx 10\, \phi^2 f_* \ , 
\end{equation}
where \(\Lambda\) is the plasma parameter (number of particles within a
Debye volume), such that
\(\ln\Lambda = 23.267 + 1.5 \ln T_4 - 0.5 \ln n\).  At highly super-thermal
speeds, the Coulomb drag falls asymptotically as
\(f\drag \propto 1 / w^{2}\).  The thermal speed of electrons is higher than
that of the protons by a factor of \((m_p / m_e)^{1/2}\), so that the
electron Coulomb drag gives a second peak fof similar strength, but at
\(w \approx \SI{430}{km.s^{-1}}\).  The behavior of \(f\drag\) in all these
different regimes is summarised in Table~\ref{tab:fdrag-regimes}, in
terms of \(\phi\) and \(w_{10} = w / \SI{10}{km.s^{-1}}\).

@@

In Appendix~\ref{sec:gas-free-bow} we calculate the behaviour of an
incoming stream of dust grains, subject only to the repulsive
radiative force from a star.  We are now in a position to see how
gas--grain drag will modify this picture.  From
equations~\eqref{eq:dust-rad-force} and~\eqref{eq:fstar}, we can write
the radiation force acting on a grain as
\begin{equation}
  \label{eq:frad-Xi}
  f\rad = \Qp\, \Xi\, f_* \ ,
\end{equation}
where \(\Qp\) is the grain's radiation pressure efficiency (see
footnote~\ref{fn:Qp} in Appendix~\ref{sec:gas-free-bow}) and \(\Xi\) is
the local radiation parameter, defined as the ratio of direct stellar
radiation pressure to gas pressure:
\begin{equation}
  \label{eq:Xi-Prad-over-Pgas}
  \Xi \equiv \frac{P\rad}{P\gas} \approx \frac{L}{4 \pi R^2 c\, (2 n k T)} \ ,
\end{equation}
where the last expression corresponds to the optically thin limit.
The grain potential \(\phi\) is also primarily determined by \(\Xi\), as
shown in Appendix~\ref{sec:cloudy-models-dust} and the inset graph of
Figure~\ref{fig:gas-grain-drag-photoionized}, but with a slow
dependence, which can be approximated\footnote{%
  There is a slight secondary dependence on the grain composition and
  stellar spectrum.  The relationship given in
  eq.~\eqref{eq:phi-vs-Xi} is appropriate for graphite grains and for
  stellar effective temperatures in the range \SIrange{20}{30}{kK}.
  For hotter stars than this, \(\phi\) should be multiplied by a further
  factor of \(1.5\), while for silicate grains it should be divided by
  \(1.5\).  } %
as
\begin{equation}
  \label{eq:phi-vs-Xi}
  \phi \approx 1.5 \bigl( 2.3 +  \ln \Xi \bigr) \ .
\end{equation}

In the outer regions of the photoionized volume around an OB star,
close to the ionization front, the radiation parameter is low, with
typical value \(\Xi \sim 0.1\).  In this regime, the negative charge
current at the grain surface due to electron collisions is roughly in
balance with the positive current due to the ultraviolet photoelectric
effect \citep{Weingartner:2001b}, leading to a low grain potential,
\(\abs{\phi} < 1\), which may be positive or negative.  The low
\(\Xi\) means that the radiative force is also weak:
\(f\rad \sim 0.1 f_*\) from equation~\eqref{eq:frad-Xi} if
\(\Qp \sim 1\) at UV wavelengths, which is true for all but the smallest
grains.  Thus, from equation~\eqref{eq:fdrag-epstein}, the radiative
force can be balanced by Epstein drag if \(w_{10} \sim 0.1\), leading to
a small drift velocity, \(w\drift \sim \SI{1}{km.s^{-1}}\), of the grains
with respect to the gas.

As a dust grain
approaches a star, the radiation parameter \(\Xi\) will increase as
\(R^{-2}\)

>>>>

As \(\abs{\phi}\) increases, long-range electrostatic interactions with
protons within the Debye radius (Coulomb drag) become increasingly
important at subsonic relative velocities, as shown in
Figure~\ref{fig:gas-grain-drag-photoionized} by the orange
(\(\phi = 1\)), green (\(\phi = 4\)), and red (\(\phi = 16\)) lines.  However,
the Coulomb drag has a peak when \(w\) is equal to the thermal
speed of the colliders, which is \(\approx \SI{10}{km.s^{-1}}\) for protons,
giving a maximum strength of
\begin{equation}
  \label{eq:fdrag-maximum}
  f_{\mathrm{max}} = 0.5\, (\ln\Lambda)\, \phi^2 f_* \approx 10\, \phi^2 f_* \ , 
\end{equation}
where \(\Lambda\) is the plasma parameter (number of particles within a
Debye volume), such that
\(\ln\Lambda = 23.267 + 1.5 \ln T_4 - 0.5 \ln n\).  At highly super-thermal
speeds, the Coulomb drag falls asymptotically as
\(f\drag \propto 1 / w^{2}\).  The thermal speed of electrons is higher than
that of the protons by a factor of \((m_p / m_e)^{1/2}\), so that the
electron Coulomb drag gives a second peak fof similar strength, but at
\(w \approx \SI{430}{km.s^{-1}}\).  The behavior of \(f\drag\) in all these
different regimes is summarised in Table~\ref{tab:fdrag-regimes}, in
terms of \(\phi\) and \(w_{10} = w / \SI{10}{km.s^{-1}}\).  This is
further illustrated in Figure~\ref{fig:drag-v-phi-plane}, where the
drag regimes are located on the \((w, \abs{\phi})\) plane.

@@

In Appendix~\ref{sec:gas-free-bow} we calculate the behaviour of an
incoming stream of dust grains, subject only to the repulsive
radiation force from a star.  For an initial inward radial trajectory,
the dust grain motion is decelerated and turned around, reaching a
minimum radius \(R\starstar\), given by equation~\eqref{eq:dust-r0}.
This drag-free radiative turnaround radius, \(R\starstar\), is smaller
for higher initial inward velocities, but is independent of the
density of the incoming stream.  We are now in a position to see how
gas--grain drag will modify this picture.

From equations~\eqref{eq:dust-rad-force} and~\eqref{eq:fstar}, we can
write the radiation force acting on a grain as
\begin{equation}
  \label{eq:frad-Xi}
  f\rad = \Qp\, \Xi\, f_* \ ,
\end{equation}
where \(\Qp\) is the grain's radiation pressure efficiency (see
footnote~\ref{fn:Qp} in Appendix~\ref{sec:gas-free-bow}) and \(\Xi\) is
the local radiation parameter, defined as the ratio of direct stellar
radiation pressure to gas pressure:
\begin{equation}
  \label{eq:Xi-Prad-over-Pgas}
  \Xi \equiv \frac{P\rad}{P\gas} \approx \frac{L}{4 \pi R^2 c\, (2 n k T)} \ ,
\end{equation}
where the last expression corresponds to the optically thin limit.
The grain potential \(\phi\) is also primarily determined by \(\Xi\), as
shown in Appendix~\ref{sec:cloudy-models-dust} and the inset graph of
Figure~\ref{fig:gas-grain-drag-photoionized}, but with a slow
dependence, which can be approximated as
\begin{equation}
  \label{eq:phi-vs-Xi}
  \phi \approx 1.5 \bigl( 2.3 +  \ln \Xi \bigr) \ .
\end{equation}
There are also slight secondary dependencies on the grain composition and
stellar spectrum.  The relationship given in eq.~\eqref{eq:phi-vs-Xi}
is appropriate for graphite grains and for stellar effective
temperatures in the range \SIrange{20}{30}{kK}.  For hotter stars than
this, \(\phi\) should be multiplied by a further factor of \(1.5\), while
for silicate grains it should be divided by \(1.5\).

In the outer regions of the photoionized volume around an OB star,
close to the ionization front, the radiation parameter is low, with
typical value \(\Xi \sim 0.1\).  In this regime, the negative charge
current at the grain surface due to electron collisions is roughly in
balance with the positive current due to the ultraviolet photoelectric
effect \citep{Weingartner:2001b}, leading to a low grain potential,
\(\abs{\phi} < 1\), which may be positive or negative.  The low
\(\Xi\) means that the radiative force is also weak:
\(f\rad \sim 0.1 f_*\) from equation~\eqref{eq:frad-Xi} if
\(\Qp \sim 1\) at UV wavelengths, which is true for all but the smallest
grains.  Thus, from the equations for \(f\drag\) given in
Table~\ref{tab:fdrag-regimes}, the radiative force can be balanced by
Epstein drag if \(w_{10} \sim 0.1\), leading to a small equilibrium drift
velocity, \(w\drift < \SI{1}{km.s^{-1}}\), of the grains with respect
to the gas.  This drift is much smaller than the inward stream
velocities that we are considering
(\(v_\infty > \SI{10}{km.s^{-1}}\)), so the dust follows the gas stream at
a slightly reduced velocity (\(< 10\%\)), and (by mass conservation) a
slightly increased density.  Each grain exerts an exactly opposite
force to \(f\drag\) upon the gas, but since the dust-gas mass ratio,
\(Z\grain\), is small, this produces a negligible acceleration of the
gas.

As the dusty stream approaches the star, the radiation parameter
\(\Xi\) will increase, with a dependence of \(R^{-2}\) once the stream
is well away from the ionization front.  This increases \(f\rad\)
(eq.~\eqref{eq:frad-Xi}), but also increases the grain potential,
\(\phi\) (eq.~\eqref{eq:phi-vs-Xi}) due to the increasing dominance of
grain charging by photoelectric ejection.  Initially, this results in
a lowering of the equilibrium drift velocity to \(w_{10} \sim 0.01\) as
the Coulomb drag kicks in (see Appendix~\ref{sec:cloudy-models-dust}).
However, at smaller radii the slow logarithmic increase in
\(\phi(\Xi)\) means that the drift velocity must start increasing again to
accommodate the linear increase of \(f\rad(\Xi)\).  Eventually,
\(f\rad\) exceeds \(f_{\mathrm{max}}\), the maximum drag force that
proton Coulomb interactions can provide
(eq.~\eqref{eq:fdrag-maximum}).  This occurs at a critical value of
the radiation parameter, which we denote the \textit{rip point}:
\(\Xi_\dag \sim 1000\).  The variations in \(\Xi_\dag\) with star and grain
parameters, which are of order \SI{+- 0.5}{dex}, are listed in
Table~\ref{tab:Xi-rip}.

\begin{table}
  \caption{Critical values of radiation parameter at the rip point: \(\Xi_\dag\)}
  \centering
  \begin{tabular}{l @{\quad\quad\quad} S S} \toprule
    & \multicolumn{2}{c}{Grain composition} \\
    Spectrum & {Graphite} & {Silicate}
    \\ \midrule
    B star & 1000 +- 400 & 350 +- 150 \\
    O star & 3000 +- 500 & 2500 +- 500 \\
    \bottomrule
    \addlinespace
    \multicolumn{3}{@{}p{0.6\columnwidth}@{}}{
    Uncertainties represent variations with grain size and gas density
    (see Appendix~\ref{sec:cloudy-models-dust}).}
  \end{tabular}
  \label{tab:Xi-rip}
\end{table}
